Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-09 paper/related.tex
Edits.

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
system and lightweight runtime checks. The type system reduces the
problem of ensuring safety of all memory accesses in a program into a
problem of enforcing a state-transition discipline on memory accesses
over a small subset of objects in the program. The later can be done
efficiently at runtime. his lets us guarantee safety in presence of
dynamic memory allocations and deallocations without requiring either
restrictive static mechanisms (e.g., linear types and borrow checker)
or expensive runtime mechanisms (e.g., garbage collection and
reference counting). Lastly, our region type system comes equipped
with full type inference that completely eliminates the need to write
region annotations on types to convince the type checker that the
program is safe.

% In contrast, ~\cite{tofte97} proposes to
% extend Standard ML, a higher-order functional language, with
% lexically-scoped stack regions. Cyclone~\cite{cyclone02}

% and defines an elaboration from
% Standard ML to the extended version of Standard ML. The aim of the
% elaboration is to introduce stack regions and do away with GC in a
% transparent fashion without jeopardizing memory safety. We too define
% an elaboration, but our focus is on introducing region annotations
% necessary to prove the safety of an object-oriented program with
% (stack and dynamic) regions. Cyclone

>>>>

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
system and lightweight runtime checks. The type system reduces the
problem of ensuring safety of all memory accesses in a program into a
problem of enforcing a state-transition discipline on memory accesses
over a small subset of objects in the program. The later can be done
efficiently at runtime. This lets us guarantee safety in presence of
dynamic memory allocations and deallocations without requiring either
restrictive static mechanisms (e.g., linear types and unique pointers)
or expensive runtime mechanisms (e.g., garbage collection and
reference counting). Lastly, our region type system comes equipped
with full type inference that completely eliminates the need to write
region annotations on types to convince the type checker that the
program is safe.

~\cite{tofte97} proposes to extend Standard ML, a higher-order
functional language, with lexically-scoped stack regions, and defines
an elaboration from Standard ML to the region-annotated version of
Standard ML. The aim of the elaboration is to introduce stack regions
and do away with GC in a transparent fashion without jeopardizing
memory safety. We too define an elaboration, but our focus is on
introducing region annotations necessary to prove the safety of an
object-oriented program that already uses (stack and dynamic) regions.
Similar to their region inference algorithm, our region type inference
algorithm can deal with region-polymorphic recursion by computing
fixed points for recursive constraints. While their inference
algorithm only ever generates equality constraints, which can be
solved via unification, our type inference algorithm also generates
partial order outlives constraints, which are required to capture
subtle relationships between lifetimes of transferable regions and
stack regions. Consequently, our constraint solving algorithm is more
sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

Cyclone~\cite{cyclone02} equips C with programmer-managed stack
regions, and a typing discipline that statically guarantees the safety
of all pointer dereferences. However, types need to be written
explicitly. While some region annotations can be inferred, type
inference is purely intraprocedural, and cannot infer region types for
functions. Later proposal~\cite{cyclone04} extends Cyclone with
dynamic regions, but has to impose unique pointer restrictions to
guarantee safety. Objects that need to be shared within data
structures are allowed to have multiple references, and safety for
such objects was ensured via runtime reference counting.

An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}. Like us, they too
assume a source language with programmer-managed memory regions, and
focus on proving safety of programs written in that language. Their
source language admits various kinds of regions in order to support
shared-memory concurrency. However, all their regions have
lexically-scoped lifetimes. In contrast, we admit regions with
dynamically determined lifetimes in order to support message-passing
concurrency. We borrow outlives relation from their formal
development, and our type system bears some similarities to their's.
However, our language also admits parametric polymorphism (generics)
and higher-order functions, whose interaction is non-trivial in
context of a region type system. On the other hand, our type system
does not admit region kinds and ownership, and is considerably simpler
than theirs. Moreover, our type system comes with a type
safety result that formalizes its guarantees with respect to a
well-defined operational semantics. Like Cyclone, ~\cite{MIT03}'s
language is explicitly typed, with some support for local type
inference. Consequently, region types for methods and classes need to
be written explicitly. In contrast, we support full type inference
that eliminates such need.

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language. Nonetheless, due to aliasing
among region names, ~\cite{HMN01} requires reference counting for
regions to ensure memory safety.
